---
title: "Getting started with invasimapr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with invasimapr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

invasimapr is an R package to assess species and trait invasiveness and site
invasibility by visualising trait dispersion and computing invasion fitness.

The package implements a network invasibility framework that decomposes
invasion fitness into three components:

- **Abiotic suitability** ($r^{(z)}_{is}$): how well the environment at
  site $s$ suits invader $i$.
- **Niche crowding** ($C^{(z)}_{is}$): how much functional overlap invader $i$
  has with the resident community at site $s$.
- **Resident competition** ($S^{(z)}_{is}$): how saturated the resident
  community is at site $s$.

## Installation

Install the development version from GitHub:

```{r install}
# install.packages("remotes")
remotes::install_github("macSands/invasimapr-dev")
```

## Workflow

The core workflow consists of eight steps, each returning an updated
`invasimapr_fit` container:
```{r workflow}
library(invasimapr)

# Step 1: Prepare inputs from occurrence, trait and environmental data
fit <- prepare_inputs(
  occ_long   = my_occurrences,
  trait_wide  = my_traits,
  env_wide    = my_environment
)

# Step 2: Simulate hypothetical invader profiles
fit <- simulate_invaders(fit, n_inv = 50, method = "columnwise")

# Step 3: Trait space and crowding
fit <- prepare_trait_space(fit)

# Step 4: Model resident abundances
fit <- model_residents(fit)

# Step 5: Learn trait-dependent sensitivities
fit <- learn_sensitivities(fit)

# Step 6: Predict invader predictors
fit <- predict_invaders(fit)

# Step 7-8: Compute invasion fitness and establishment probability
fit <- predict_establishment(fit, option = "C")

# Summarise invasiveness and invasibility
fit <- summarise_results(fit)
```

## Key functions

| Function | Purpose |
|---|---|
| `prepare_inputs()` | Assemble community, trait and environmental matrices |
| `simulate_invaders()` | Generate hypothetical invader trait profiles |
| `prepare_trait_space()` | Compute trait space, convex hull, and crowding |
| `model_residents()` | Fit GLMMs to resident abundance |
| `learn_sensitivities()` | Estimate trait-dependent sensitivity slopes |
| `predict_invaders()` | Project invaders into the resident model space |
| `predict_establishment()` | Compute invasion fitness and map to probability |
| `summarise_results()` | Summarise invasiveness and invasibility |

## Further reading

For detailed tutorials with real data, see the package articles:

- [Introduction](../articles/introduction.html)
- [Step-by-step Workflow](../articles/step-by-step-workflow.html)
- [Clustering and risk scenarios](../articles/clustering-and-risk-scenarios.html)
- [Invasion fitness synthesis](../articles/invasion-fitness-synthesis.html)
- [Computing invasion fitness](../articles/computing-invasion-fitness.html)

## Acknowledgments

This software was developed with funding from the European Union's Horizon
Europe Research and Innovation Programme under grant agreement ID No
[101059592](https://doi.org/10.3030/101059592).
